{{< include _productivity-setup.md >}}

---
title: Trends in Productivity
---

### The Market and the Non-Market

Productivity is measured across the industries that make up the economy. These industries can be described as *market* and *non-market* sectors. The market sector includes industries that produce goods and services which are sold at market prices. That means the price is reflective of the demand. The non-market sector covers education and training, health care and social assistance, and public administration and safety. Employment growth has been much faster in the non-market sector, as shown in @fig-employment-market-non-market below.

```{r}
#| label: fig-employment-market-non-market
#| fig-cap: Share of employment in the market and non-market sectors

anzsic_codes <- anzsic2006 |> 
  select(contains("_division")) |> 
  distinct()

industry_employment <- read_abs_local("6291.0.55.001", path = "data") 

sector_employment <- industry_employment |> 
  filter(series_type == "Seasonally Adjusted") |> 
  separate_series(c("anzsic_division", "indicator"),
                  remove_nas = T) |> 
  inner_join(anzsic_codes) |> 
  mutate(sector = case_when(
    anzsic_division_code %in% LETTERS[c(1:14, 18, 19)] ~ "Market Sector", 
    anzsic_division_code %in% c("O", "P", "Q") ~ "Non-Market Sector"
  )) |> 
  group_by(date, sector) |> 
  summarise(value = sum(value),
            .groups = "drop") |> 
  group_by(date) |> 
  mutate(share = value/sum(value), .groups = "drop") |> 
  filter(sector == "Non-Market Sector")

ggplot(sector_employment, aes(x = date, y = share, col = sector)) + 
  geom_line() +
  theme_nrc() +
  labs(title = "Share of Employment in the Non-Market Sector", 
       subtitle = "Seasonally Adjusted; Quarterly",
       x = NULL,
       y = NULL) + 
  scale_y_continuous(labels = label_percent()) +
  scale_x_date(breaks = plot_date_range(sector_employment),
               date_labels = "%e %b\n%Y")
```

Estimates of productivity in the non-market sector should be interpreted with caution. What is the unit of output that is being measured? For the education sector, output might be the number of students who graduate. In that case, measured productivity would increase by reducing the number of teachers - but that would reduce the quality of education and result in worse outcomes for society.

## Trends in Productivity

Productivity across the market and non-market sectors, as well as the total productivity across the Australian economy, is shown in @fig-productivity. Productivity growth began to slow from from 2015, although the pandemic created a productivity bubble. Recent declines in productivity growth may simply be a return to normal.

```{r}
#| label: fig-productivity
#| fig-cap: Long run productivity, market, non-market, and total
national_accounts_productivity <- national_accounts |> 
 filter(series %in% c("GDP per hour worked: Index ;",
                       "Gross value added per hour worked market sector: Index ;",
                       "Gross value added per hour worked non-market sector: Index ;"),
        series_type == "Seasonally Adjusted",
        !is.na(value)) |> 
  mutate(series = case_when(
           series == "GDP per hour worked: Index ;" ~ "Labour Productivity",
           series == "Gross value added per hour worked market sector: Index ;" ~ "Labour Productivity (Market)",
           series == "Gross value added per hour worked non-market sector: Index ;" ~ "Labour Productivity (Non-Market)"
         )) 

national_accounts_productivity |> 
  ggplot(aes(x = date, y = value, col = series)) +
  geom_line() +
  scale_colour_nrc() +
  scale_x_date(breaks = plot_date_range(national_accounts_productivity),
               date_labels = "%e %b\n%Y") +
  labs(title = "GDP per Hour Worked (B YYYY = 100)",
       subtitle = "Seasonally Adjusted; Quarterly") +
  theme_nrc(legend = 'bottom') +
  labs(y = NULL,
       x = NULL)
```

The 20-year average of productivity growth confirms this trend, as shown in @fig-productivity-cagr.

```{r}
#| label: fig-productivity-cagr
#| fig-cap: Productivity growth, 20 year average compound growth rate. 
national_accounts_20 <- national_accounts|> 
  filter(series %in% c("GDP per hour worked: Index ;"),
         series_type == "Seasonally Adjusted") |> 
  mutate(cagr = (value/lag(value, n = 80))^(1/80) - 1) |> 
  filter(!is.na(cagr)) 

national_accounts_20 |> 
  ggplot(aes(x = date, y = cagr, col = series)) +
  geom_line() + 
  scale_y_continuous(labels = scales::label_percent()) +
  scale_x_date(breaks = plot_date_range(national_accounts_20),
               date_labels = "%e %b\n%Y") +
  labs(title = "Long Run Productivity Growth",
       subtitle = "20 year average compound growth rate",
       y = NULL,
      x = NULL) + 
  theme_nrc()
```

Because productivity is measured as GDP per hour worked, it is important to look at what is happening with measures of GDP and hours worked.

### Trends in Hours Worked

@fig-hours shows the long run trend for hours worked in the market and non-market sectors, and aggregated across the economy. There is a clear increase in the growth of hours worked across all sectors of the economy, but especially in the non-market sector. In fact the growth of hours worked in the non-market sector has been three times faster since Covid than before.

Increasing hours worked is a drag on productivity. One of the main reasons that people choose to work more hours is that their wage is not sufficient to meet their material needs.

```{r}
#| label: fig-hours
#| fig-cap: "Hours Worked: Index (June 2022 = 100)"
national_accounts_hw <- national_accounts |> 
  filter(series %in% c("Hours worked: Index ;",
                       "Hours worked market sector: Index ;",
                       "Hours worked non-market sector: Index ;"),
         series_type == "Seasonally Adjusted",
         !is.na(value)) |> 
  mutate(series = case_when(
    series == "Hours worked: Index ;" ~ "Hours Worked",
    series == "Hours worked market sector: Index ;" ~ "Hours Worked (Market)",
    series == "Hours worked non-market sector: Index ;" ~ "Hours Worked (Non-Market)"
  ))

national_accounts_hw |> 
  ggplot(aes(x = date, y = value, col = series)) + 
  scale_x_date(date_labels = "%e %b\n %Y",
               breaks = plot_date_range(national_accounts_hw)) +
  scale_colour_nrc() +
  geom_line() + 
  theme_nrc(legend = "bottom") + 
  labs(x = NULL, 
       y = NULL,
       title = "Hours Worked: Index (June 2022 = 100)",
       subtitle = "Seasonally Adjusted; Quarterly")
```

### Trends in GDP

On the other side of the equation is GDP, with the long run trend shown in @fig-gdp. While the long run growth prior to the pandemic was slightly stronger than over the past five years, in comparison to hours worked, the impact on productivity is minimal.

```{r}
#| label: fig-gdp
#| fig-cap: Gross Domestic Product
national_accounts_gdp <- national_accounts |> 
  filter(series == "Gross domestic product: Chain volume measures ;",
         series_type == "Seasonally Adjusted",
         !is.na(value)) 

national_accounts_gdp |> 
  ggplot(aes(x = date, y = value, col = 'hours')) + 
  geom_line() + 
  scale_x_date(date_labels = "%e %b\n %Y",
               breaks = plot_date_range(national_accounts_gdp)) +
  theme_nrc() + 
  labs(x = NULL,
       y = NULL, 
       title = "Gross Domestic Product",
       subtitle = "Seasonally Adjusted; Quarterly") + 
  scale_y_continuous(labels = scales::label_currency(scale = 1e-3, suffix = "bn"))
```
