{{< include _inflation-setup.md >}}

---
title: Inflation and Wages
---

# Inflation and Wages

When discussing inflation, we need to consider the relationship between increasing prices and increasing wages. If wages are increasing at the same rate as prices, then no ones purchasing power is getting any better or worse. How this feels will depend on the historical relationship between prices and wages.

The Wage Price Index (**WPI**) measures the change in the price of wages. The real value of wages can be calculated by looking at the difference between the WPI, and the CPI, as shown in @fig-wpi-cpi. This shows that on average, prices have increased by 25.1%, and wages have increased by 22.1%. So in real terms, the purchasing power of the average wage has fallen by 3% - a decline in standard of living.

```{r}
#| label: fig-wpi-cpi
#| fig-cap: Change in real value of wages since March 2020. 
cumulative_cpi <- cpi_quarterly |> 
  filter(indicator == "Index Numbers",
         state == "Australia",
         cpi_expenditure_class == "All groups CPI",
         date >= "2020-03-01") |> 
  mutate(cumulative_inflation = value - value[date=="2020-03-01"]) |> 
  select(date, cumulative_inflation)

cumulative_wpi <- wpi_quarterly |> 
  filter(state == "Australia",
         unit == "Index Numbers",
         indicator == "Total hourly rates of pay excluding bonuses",
         sector == "Private and Public",
         industry == "All industries",
         date >= "2020-03-01") |> 
  mutate(cumulative_wages = value - value[date=="2020-03-01"]) |> 
  select(date, cumulative_wages) |> 
  distinct()


cumulative <- inner_join(cumulative_wpi, cumulative_cpi) |> 
  mutate(cumulative_real_wage = cumulative_wages - cumulative_inflation) |> 
  pivot_longer(cols = -date,
               values_to = "value",
               names_to = "indicator") |> 
  mutate(indicator = case_when(
    indicator == "cumulative_wages" ~ "Hourly rates of pay",
    indicator == "cumulative_inflation" ~ "CPI",
    indicator == "cumulative_real_wage" ~ "Real hourly rates of pay"
  )) |> 
  mutate(value_label = ifelse(date == max(date),
                              scales::label_percent(scale = 1, accuracy = 0.1)(value),
                              NA))

cumulative |> 
  ggplot(aes(x = date, y = value, colour = indicator)) +
  geom_line() +
  scale_x_date(breaks = plot_date_range(cumulative),
               date_labels = "%e %b\n %Y") +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  scale_colour_nrc() +
  labs(x = NULL,
       y = NULL,
       title = "Wages Have Not Kept Up With Inflation",
       subtitle = "Cumulative change in hourly rates of pay and the CPI since March 2020") +
  geom_dl(method = "last.qp", aes(label = value_label)) + 
  theme_nrc(legend = "bottom")

```

What would this chart look like if the average worker recieved a 6% pay rise for the next year, and inflation grew at the RBA's target of 2.5%?. @fig-wpi-cpi-fc shows that this would only represent breaking even with the real value of wages in March 2020. We're a long far behind at ***improving***living standards.

```{r}
#| label: fig-wpi-cpi-fc
#| fig-cap: Forecast change in real value of wages since March 2020
fc <- tibble(
  date = as_date(c("2025-09-01", "2025-12-01", "2026-03-01", "2026-06-01")),
  cumulative_wages = rep(22.1+6, 4),
  cumulative_inflation = c(25.1+2.5/4, 25.1+2.5/3, 25.1+2.5/2, 25.1+2.5)
) |> 
  mutate(cumulative_real_wage = cumulative_wages - cumulative_inflation)

cumulative_forecast <- inner_join(cumulative_wpi, cumulative_cpi) |> 
  mutate(cumulative_real_wage = cumulative_wages - cumulative_inflation) |> 
  bind_rows(fc) |> 
  pivot_longer(cols = -date,
               values_to = "value",
               names_to = "indicator") |> 
  mutate(indicator = case_when(
    indicator == "cumulative_wages" ~ "Hourly rates of pay",
    indicator == "cumulative_inflation" ~ "CPI",
    indicator == "cumulative_real_wage" ~ "Real hourly rates of pay"
  )) |> 
  mutate(value_label = ifelse(date == max(date),
                              scales::label_percent(scale = 1, accuracy = 0.1)(value),
                              NA))

cumulative_forecast |> 
  ggplot(aes(x = date, y = value, colour = indicator)) +
  geom_line() +
  scale_x_date(breaks = plot_date_range(cumulative),
               date_labels = "%e %b\n %Y") +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  scale_colour_nrc() +
  labs(x = NULL,
       y = NULL,
       title = "Wages Have Not Kept Up With Inflation",
       subtitle = "Cumulative change in hourly rates of pay and the CPI since March 2020 with 6% pay rise for 2025-26 and 2.5% inflation ") +
  geom_dl(method = "last.qp", aes(label = value_label)) + 
  theme_nrc(legend = "bottom")
  
```

Different economists will use different prices to convert nominal dollars into real dollars. The one they choose will often be the one that tells the story they want to tell. @fig-wpi-cost-of-living shows a hypothetical the change in the real value of an annual salary of \$45,000 between March 1999 and today, under three scenarios - wages offset by changes in consumer price index, wages offset by changes in the cost of living, and wages increased in real terms by 3% each year.

```{r}
#| label: fig-wpi-cost-of-living
#| fig-cap: "Change in real value of an annual wage of $45,000 by comparison with CPI and LCI"
cpi_pc <- read_abs("6401.0", tables = "1", retain_files = TRUE, path = "data") |> 
  filter(series == "Percentage Change from Corresponding Quarter of Previous Year ;  All groups CPI ;  Australia ;",
         series_type == "Original") |> 
  select(date, cpi_pc = value)

lci_pc <- lci |> 
  filter(data_type == "Percentage Change from Corresponding Quarter of Previous Year",
         cpi_category == "All groups",
         household_type == "Employee households") |> 
  select(date, lci_pc = value)

wpi_pc <-  wpi_quarterly |> 
  filter(indicator == "Total hourly rates of pay excluding bonuses",
         data_type == "Quarterly Index",
         state == "Australia",
         sector == "Private and Public",
         industry == "All industries") |> 
  distinct() |> 
  select(date, wpi = value) |> 
  mutate(wpi_pc = 100*(wpi-lag(wpi, 4))/lag(wpi, 4))

wage_price <- wpi_pc |> 
  inner_join(cpi_pc) |>
  inner_join(lci_pc) |> 
  filter(!is.na(lci_pc)) |> 
  mutate(real_wage_growth = wpi_pc - cpi_pc,
         real_wage_growth_cost_of_living = wpi_pc - lci_pc,
         real_wage_constant = 3/4) |> 
  filter(!is.na(wpi_pc)) |> 
  mutate(real_wage = 45000+(45000*cumsum(real_wage_growth)/100),
         real_wage_3 = 45000+(45000*cumsum(real_wage_constant/100)),
         real_wage_lci = 45000+(45000*cumsum(real_wage_growth_cost_of_living)/100)) |> 
  add_row(.before = 1, 
          date = as_date("1999-03-01"),
          real_wage = 45000,
          real_wage_3 = 45000,
          real_wage_lci = 45000)

wage_price |> 
  pivot_longer(cols = c(real_wage, real_wage_lci, real_wage_3)) |> 
  mutate(name = case_when(
    name == "real_wage" ~ "Real Wage (CPI)",
    name == "real_wage_lci" ~ "Real Wage (Cost of Living)",
    name == "real_wage_3" ~ "Constant 3% Real Wage Growth",
  )) |> 
  ggplot(aes(x = date, y = value, col = name)) + 
  geom_line() +
  theme_nrc(legend = 'bottom') +
  scale_colour_nrc() + 
  labs(x = NULL,
       y = NULL,
       title = "Value of real wages that were equal to $45,000 in March 1999") +
  scale_y_continuous(labels = scales::label_currency())
```
